Explaining Black-Box Models through Counterfactuals

### 2023-11-15 examples/CounterfactualExplanations.jl/paper.qmd
ce done

A large and growing number of counterfactual [generators](https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/explanation/generators/overview/) have already been implemented in our package (@tab-gen). At a high level, we distinguish generators in terms of their compatible model types, their default search space, and their composability. All "gradient-based" generators are compatible with differentiable models, e.g. `Flux` and `torch`, while "tree-based" generators are only applicable to models that involve decision trees. Concerning the search space, it is possible to search counterfactuals in a lower-dimensional latent embedding of the feature space that implicitly encodes the data-generating process (DGP). To learn the latent embedding, existing work has typically relied on generative models or existing causal knowledge [@joshi2019realistic, @karimi2021algorithmic]. While this notion is compatible with all of our gradient-based generators, only some generators search a latent space by default. Finally, composability implies that the given generator can be blended with any other composable generator, which we discuss in @sec-gen-comp.

@@

::: {#tab-gen}

```{=latex}
\begin{tabular}{llll}
\toprule
Generator & Model Type & Search Space & Composable\\
\midrule
ClaPROAR [@altmeyer2023endogenous] & gradient based & feature & yes\\
CLUE [@antoran2020getting] & gradient based & latent & yes\\
DiCE [@mothilal2020explaining] & gradient based & feature & yes\\
FeatureTweak [@tolomei2017interpretable] & tree based & feature & no\\
Gravitational [@altmeyer2023endogenous] & gradient based & feature & yes\\
Greedy [@schut2021generating] & gradient based & feature & yes\\
GrowingSpheres [@laugel2017inversea] & agnostic & feature & no\\
PROBE [@pawelczyk2022probabilistically] & gradient based & feature & no\\
REVISE [@joshi2019realistic] & gradient based & latent & yes\\
Wachter [@wachter2017counterfactual] & gradient based & feature & yes\\
\bottomrule
\end{tabular}
```

Overview of implemented counterfactual generators.

:::

>>>>

A large and growing number of counterfactual [generators](https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/explanation/generators/overview/) have already been implemented in our package (@tbl-gen). At a high level, we distinguish generators in terms of their compatible model types, their default search space, and their composability. All "gradient-based" generators are compatible with differentiable models, e.g. `Flux` and `torch`, while "tree-based" generators are only applicable to models that involve decision trees. Concerning the search space, it is possible to search counterfactuals in a lower-dimensional latent embedding of the feature space that implicitly encodes the data-generating process (DGP). To learn the latent embedding, existing work has typically relied on generative models or existing causal knowledge [@joshi2019realistic, @karimi2021algorithmic]. While this notion is compatible with all of our gradient-based generators, only some generators search a latent space by default. Finally, composability implies that the given generator can be blended with any other composable generator, which we discuss in @sec-gen-comp.

@@

\onecolumn

| Generator | Model Type | Search Space | Composable |
|---|---|---|---|
| ClaPROAR [@altmeyer2023endogenous] | gradient based | feature | yes |
| CLUE [@antoran2020getting] | gradient based | latent | yes |
| DiCE [@mothilal2020explaining] | gradient based | feature | yes |
| FeatureTweak [@tolomei2017interpretable] | tree based | feature | no |
| Gravitational [@altmeyer2023endogenous] | gradient based | feature | yes |
| Greedy [@schut2021generating] | gradient based | feature | yes |
| GrowingSpheres [@laugel2017inversea] | agnostic | feature | no |
| PROBE [@pawelczyk2022probabilistically] | gradient based | feature | no |
| REVISE [@joshi2019realistic] | gradient based | latent | yes |
| Wachter [@wachter2017counterfactual] | gradient based | feature | yes |

: Overview of implemented counterfactual generators. {#tbl-gen}

\twocolumn
